---
title: "Class 11: Alphafold"
author: "Julia Di Silvestri (A16950824)"
format: pdf
editor: visual
---

AlphaFold has changed the game for protein structure prediction and allows anyone with sufficient bioinformatics skills to predict the strucutre of virtually any protein.

We ran AlphaFold via GoogleColab at:
https://colab.research.google.com/github/sokrypton/ColabFold/blob/main/AlphaFold2.ipynb

In particular we used their AlphaFold2_mmseqs2 version that uses mmseq2 rather than HMMer for sequence search

The main outputs include a set of **PDB structure** files along with matching **JSON format files** that tell us how good the resulting models might be

Let's start by loading these structures up in Mol*

```{r}
library(bio3d)

results_dir <- "hivdimer_23119_0"

# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

```{r}
# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
pdbs
```

```{r}
rd <- rmsd(pdbs, fit=T)

range(rd)
```

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

Finding the most consistent "core" common across all the models:

```{r}
core <- core.find(pdbs)

#write a directory for the fitte structures
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

Now we can examine the RMSF between positions of the structure. 

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

If the predicted model has more than one domain, each domain may have high confidence, yet the relative positions of the domains may not. 

## Predicted Alignment Error for Domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```




